---
title: "Difference-in-Differences: Organ Donation Rates"
format: html
---

## Background

In Q3 2011, California began requiring residents to answer Yes or No about organ
donation when renewing their driver's license (an "active choice" policy). Other
states did not change their policy. We can use **Difference-in-Differences (DiD)**
to estimate whether the policy caused a change in donation rates, using other
states as controls.

**Data source:** Kessler, Judd B. and Alvin E. Roth. 2014. "Don't Take 'No' For
An Answer: An Experiment With Actual Organ Donor Registrations." NBER Working
Paper 20378. Dataset via the `causaldata` R package, featured in *The Effect* by
Nick Huntington-Klein.

The dataset is in `data/organ_donations.csv` with columns: `State`, `Quarter`,
`Rate`, `Quarter_Num`. It covers 27 states (including DC) across 6 quarters
(Q4 2010 through Q1 2012).


# Phase 1: Concept & Idea Generation

Use Claude Code to explore the research question before writing any code. Open a
terminal and run `claude`, then try prompts like:

- *What is Difference-in-Differences and when is it an appropriate method?*
- *What assumptions does DiD rely on? How could they be violated?*
- *I have organ donation rate data for 27 states across 6 quarters. California
  introduced an "active choice" policy in Q3 2011. How would I set up a DiD
  analysis to estimate the policy's effect?*


# Phase 2: Script Creation

Ask Claude Code to help you write code to import the dataset and prepare it for
analysis. For example: *Read in data/organ_donations.csv and add the variables I
need for a DiD analysis: a treatment indicator for California and a post-policy
indicator for Q3 2011 onward.*

```{r}
#| label: setup
library(tidyverse)
library(broom)
```

```{r}
#| label: import-and-prepare
# Use Claude Code to help you write this section.
# The code in Phase 3 expects these variable names:
#   organ          = the dataframe
#   treated_state  = factor: "California" or "Control"
#   post           = logical: TRUE when Quarter_Num >= 4

```


# Phase 3: Analysis & Visualization

The code below runs the DiD analysis and creates a plot. Use Claude Code to
explore and modify it. Try changing the plot style, adjusting scales, or adding
new analyses.

## Visualize parallel trends

```{r}
#| label: plot-trends
quarter_labels <- c(
  "1" = "Q4\n2010", "2" = "Q1\n2011", "3" = "Q2\n2011",
  "4" = "Q3\n2011", "5" = "Q4\n2011", "6" = "Q1\n2012"
)

group_means <- organ |>
  summarise(
    mean_rate = mean(Rate),
    .by = c(treated_state, Quarter, Quarter_Num)
  )

palette <- c("California" = "#D55E00", "Control" = "#0072B2")

ggplot(group_means, aes(
  x = Quarter_Num, y = mean_rate,
  colour = treated_state, shape = treated_state
)) +
  annotate(
    "rect",
    xmin = 3.5, xmax = 6.5, ymin = -Inf, ymax = Inf,
    fill = "grey90", alpha = 0.5
  ) +
  geom_vline(xintercept = 3.5, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1) +
  geom_point(size = 3.5) +
  annotate(
    "text",
    x = 3.4, y = max(group_means$mean_rate) + 0.005,
    label = "Policy\nenacted",
    hjust = 1, size = 3.2, colour = "grey40", lineheight = 0.9
  ) +
  annotate(
    "label",
    x = 5, y = min(group_means$mean_rate) - 0.01,
    label = "Post-treatment period",
    size = 3, fill = "grey90", label.padding = unit(0.2, "lines"),
    colour = "grey40"
  ) +
  scale_x_continuous(breaks = 1:6, labels = quarter_labels) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = palette) +
  scale_shape_manual(values = c("California" = 17, "Control" = 16)) +
  labs(
    title = "Organ Donation Rates: California vs. Control States",
    subtitle = str_wrap(paste(
      "Difference-in-Differences design around California's Q3 2011",
      "active-choice organ donation policy."
    ), width = 80),
    x = NULL, y = "Donation Rate",
    colour = NULL, shape = NULL,
    caption = "Source: Kessler & Roth (2014) via causaldata R package"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle = element_text(colour = "grey30", margin = margin(b = 10)),
    plot.caption = element_text(colour = "grey50", size = 9),
    legend.position = "top",
    legend.justification = "left",
    panel.grid.minor = element_blank()
  )
```

## Estimate the DiD model

```{r}
#| label: did-model
did_model <- lm(Rate ~ treated_state * post, data = organ)
tidy(did_model, conf.int = TRUE)
```

Try asking Claude Code:

- *Change the y-axis to show values from 0% to 100%*
- *Switch the plot to a black-and-white theme suitable for print*
- *Add robust standard errors clustered at the state level*
- *Create a summary statistics table broken down by treatment group and period*


# Phase 4: Debugging

The code below has several bugs. Use Claude Code to find and fix them. Try
copying a chunk into your prompt, or ask Claude Code to *look at the Phase 4
section of example.qmd and fix the bugs*.

```{r}
#| label: buggy-imports
#| eval: false
# Bug 1: Wrong file path and missing library
organ_data <- read_csv("organ_donations.csv")

glimpse(organ_data)
```

```{r}
#| label: buggy-variables
#| eval: false
# Bug 2: Wrong column names and logic error
organ_data <- organ_data |>
  mutate(
    treated = if_else(state == "CA", "Treated", "Control"),
    post_policy = quarter_num > 4
  )
```

```{r}
#| label: buggy-plot
#| eval: false
# Bug 3: Mismatched variable names and wrong geom
ggplot(organ_data, aes(x = Quarter_Num, y = rate, colour = treated)) +
  geom_bar() +
  labs(title = Organ Donation Rates Over Time)
```

```{r}
#| label: buggy-model
#| eval: false
# Bug 4: Formula references wrong variable names
model <- lm(rate ~ treated * post_policy, data = organ_data)
summary(model)
```
